A Plasma Torus Around a Young Low-Mass Star

### 2025-03-21 paper/ms.tex
update ms

The role of the star's magnetic field could be better understood through
new observations, and new theory.  From the theoretical perspective,
there is an urgent need for rigid-field (magneto)-hydrodynamic modeling
to go beyond previous work \cite{Townsend2005,Townsend2008} and to
explore the effects of non-dipolar field contributions.  In particular,
dynamo simulations of fully-convective M dwarfs have suggested that
global-scale mean fields might be confined to a single hemisphere
\cite{Brown2020}; such fields could yield accumulation surfaces quite
different from those that have previously been explored
\cite{Townsend2008}.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent probe could also be connected the recent work
\cite{Kaur2024} showing that CPVs are variable radio emitters with
emission components that can be not only persistent, but also
short-lived and highly polarized.  In particular, detecting radio
emission produced by an electron cyclotron maser instability
(e.g. \cite{Callingham2021}) would provide a measurement of the field
strength at the site of the emitting region, further clarifying the
magnetic field's structure.

@@

{\it Effective temperature, radius, mass, and spectral classification}---We adopted
the star's effective temperature and radius measured using the spectral
energy distribution (SED) fitting procedure described by
\cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we performed the fit using the broadband magnitudes from
Gaia DR2, APASS, 2MASS, SDSS, and WISE $W1$ and $W2$.  The resulting
best-fit SED is shown in Figure~\ref{fig:sed}.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
the HIRES spectra against the photometric standards tabulated by
\cite{Bochanski2007}.   This result agrees with the effective
temperature found from the SED fitting \cite{Pecaut2013}.
We measured the equivalent width of the H$\alpha$ line by fitting a
range of models to the time-averaged line profile shown in
Figure~\ref{fig:spec}, selecting the model that minimized the Bayesian
information criterion (a sum of two Gaussians), and numerically
integrating it.
Our quoted result, EW$_{{\rm H}\alpha}$=7.2$\pm$0.2\,\AA, comes from
numerically integrating within $|\Delta v/v_{\rm eq}|<1$; 
integrating the entire line profile (including the broad wings) would yield
EW$_{{\rm H}\alpha}$=10.2$\pm$0.3\,\AA
Either value would classify the star as a ``weak-lined'' T Tauri \cite{Briceno2019}.

@@

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figures~\ref{fig:spec}
and~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination (in which Lyman-series
photons are trapped by high optical depth, so that recombinations
cascade through H$\alpha$ and other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with 
densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$ and 
temperatures between 1,000-10,000\,K,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$ suggest
$n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
We have assumed a uniform density clump, and so this yields a total
gas mass $M_{\rm gas}\approx 8\times10^{16}$\,g.

>>>>

The role of the star's magnetic field could be better understood through
new observations, and new theory.  From the theoretical perspective,
there is an urgent need for rigid-field magneto-hydrodynamic modeling to
go beyond previous work \cite{Townsend2005,Townsend2008,Krticka2022} and
to explore what field topologies might explain the observed light curve
morphologies \cite{Bouma2024}.  In particular, dynamo simulations of
fully-convective M dwarfs have suggested that global-scale mean fields
might be confined to a single hemisphere \cite{Brown2020}; such fields
could yield accumulation surfaces quite different from those that have
previously been explored.  Observationally, spectropolarimetry has the
potential to assess both the field strength and topology.  An
independent probe could also be connected the recent work
\cite{Kaur2024} showing that CPVs are variable radio emitters with
emission components that can be not only persistent, but also
short-lived and highly polarized.  In particular, detecting radio
emission produced by an electron cyclotron maser instability (e.g.
\cite{Callingham2021}) would provide a measurement of the field strength
at the site of the emitting region, further clarifying the magnetic
field's structure.

@@

{\it Effective temperature, radius, mass, and spectral classification}---We adopted
the star's effective temperature and radius measured using the spectral
energy distribution (SED) fitting procedure described by
\cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we performed the fit using the broadband magnitudes from
Gaia DR2, APASS, 2MASS, SDSS, and WISE $W1$ and $W2$.  The resulting
best-fit SED is shown in Figure~\ref{fig:sed}.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
the HIRES spectra against the photometric standards tabulated by
\cite{Bochanski2007}.   This result agrees with the effective
temperature found from the SED fitting \cite{Pecaut2013}.
We measured the equivalent width of the H$\alpha$ line by fitting a
range of models to the time-averaged line profile shown in
Figure~\ref{fig:spec}, selecting the model that minimized the Bayesian
information criterion (we found a sum of two Gaussians to be preferred), and numerically
integrating this best fit model.
Our quoted result, EW$_{{\rm H}\alpha}$=7.2$\pm$0.2\,\AA, comes from
numerically integrating within $|\Delta v/v_{\rm eq}|<1$; 
integrating over the entire line profile (including the broad wings) would yield
EW$_{{\rm H}\alpha}$=10.2$\pm$0.3\,\AA
Either value would classify the star as a weak-lined T Tauri \cite{Briceno2019}.

@@

Circumstellar H$\alpha$ emission might be sourced either from
scattering of stellar H$\alpha$ photons, or from radiative
recombination.  We neglect scattering because Figures~\ref{fig:spec}
and~\ref{fig:hirescuts} show the amplitude of the circumstellar
H$\alpha$ emission varying by a factor of $\approx$5, in a manner
uncorrelated with any variability in the chromospheric line core.  The
volume emissivity under case B recombination (in which Lyman-series
photons are trapped by high optical depth, so that recombinations
cascade through H$\alpha$ and other Balmer lines) can be written as
\begin{equation}
  j_{\rm H\alpha} = n_{\rm e} n_{\rm p} \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha},
\end{equation}
where $n_{\rm e}$ and $n_{\rm p}$ are the electron and proton densities,
and $\alpha^{\rm eff}_{\rm H\alpha}(T)$ is the effective recombination
coefficient, i.e.\ the number of recombinations per unit volume, per
unit time.
For hydrogen with 
densities $n_{\rm H}$$\sim$10$^{10}$\,cm$^{-3}$ and 
temperatures between 1,000-10,000\,K,
$\alpha^{\rm eff}_{\rm H\alpha}$ is typically on the order of $10^{-12}$ to
$10^{-13}$\,cm$^3$\,s$^{-1}$ \cite{Hummer1987,Draine2011}.
Neglecting the effects of atoms other than hydrogen, we can assume
an ionization fraction $x$, such that $n_{\rm e} = n_{\rm p} = x
n_{\rm H}$, for $n_{\rm H}$ the hydrogen number density.  Let $L_{\rm
H\alpha} = j_{\rm H\alpha} V$, for $V$ the volume of the emitting
hydrogen.
The luminosity of circumstellar hydrogen emission, $L_{\rm H\alpha}$,
is an observable: our SED fitting routine yields
$L_\star$$\approx$0.012\,$L_\odot$, which gives
$L_{\rm H\alpha},\star$$\approx$1.0$\times$$10^{28}$erg\,s$^{-1}$, and
the luminosity of the clump $L_{\rm H\alpha}$ is of order one tenth
that of the star.
Assuming an emitting sphere of radius $r$, we can write
\begin{equation}
  n_H = \frac{1}{x}
  \left( 
    \frac{L_{\rm H\alpha}}{ \alpha^{\rm eff}_{\rm H\alpha}(T) h \nu_{\rm H\alpha} V }
  \right)^{1/2}.
\end{equation}
Nominal parameters of $r = 0.1 R_\star$, $x=1/2$, and
$\alpha^{\rm eff}_{\rm H\alpha}$=$10^{-13}$\,cm$^3$\,s$^{-1}$ suggest
$n_{\rm H}\approx 5\times10^{11}$\,cm$^{-3}$.
We have assumed a uniform density clump, and so this yields a total
gas mass $M_{\rm gas}\approx 8\times10^{16}$\,g.
